Bayesian model in JAGS

### 2023-01-18 JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
add more materials

In this tutorial, we will learn how to use JAGS (Just Another Gibbs Sampler) in R to estimate unknown parameters with Markov Chain Monte Carlo (MCMC) methods in Bayesian framework. If you are interested in learning more about the theoretical knowledge of MCMC and Bayesian modeling, please check out courses "Probability and Statistics" (taught by Prof. Joseph Chang) and "Bayesian Statistics" (taught by Prof. Josh Warren). 
\
\
In the first section, we will introduce a simple jags model to help understand the syntax for rjags package.
\
\
In the second section, we will learn how to build up complicated models step by step (censor data and spatial autocorrelation). Before reading this section, you will need to familiar yourself with the RSV_timing_explain tutorials and a research article on the [relative RSV timing](https://doi.org/10.1111/irv.12965)

@@

After we estimated the [RSV timing](https://doi.org/10.1111/irv.12965) using second derivative method (please check: RSV_timing_explain), we would like to know if some potential drivers contribute to the relative timing difference in RSV epidemic onsets before and during COVID-19 pandemic.

The model is given as: 
$$Y_i=\alpha_{0i}+\beta_1\overline{Z_i}+\epsilon_i$$
$$\alpha_{0i}=\beta_0+\boldsymbol{x_i^T \lambda}+\theta_i$$

@@

where $Y_i$ is the relative timing of RSV epidemic onset with respect to Florida in 2021 in state $i$ (measured in weeks), and $\overline{Z_i}$ is the average relative timing of RSV seasons in state $i$ compared to Florida from 2016-2019. The global intercept parameter $\beta_0$ and slope parameter $\beta_1$ describe the overall similarity between a typical year’s relative onset timing across all states and the relative timings observed in 2021. The observation-level random effect $\epsilon_i  ~ N(0,\sigma^2)$ accounts for residual variability in the data.
\
\
The state-specific intercepts, $\alpha_{0i}=\beta_0+\boldsymbol{x_i^T \lambda}+\theta_i$, represent the state-level similarity in relative timing for 2021 compared to previous seasons, where $x_i$ is the vector of covariates that could potentially explain differences in the relative timing of RSV between 2021 and previous years. The covariates include population density, average household size, and stringency index of non-pharmaceutical interventions against COVID-19. 
\
\
$\theta_i$ represents a spatially correlated random effect, which we will learn more in detail later.

## Corresponding model structure without considering spatial autocorrelation

```{r}
Nospatial_model <- "model{
for(i in 1:n){
Z[i] ~ dnorm(mu[i], inv_var)
 mu[i] <- (beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i])+(beta1+x1[i]*theta1+x2[i]*theta2+theta3*x3[i])*log_W[i]
}

beta0~dnorm(0,0.0001)
beta1~dnorm(0,0.0001)
lambda1~dnorm(0,0.0001)
lambda2~dnorm(0,0.0001)
lambda3~dnorm(0,0.0001)
theta2~dnorm(0,0.0001)
theta1~dnorm(0,0.0001)
theta3~dnorm(0,0.0001)

inv_var ~ dgamma(0.01, 0.01)
}
"
```

>>>>

In this tutorial, we will learn how to use JAGS (Just Another Gibbs Sampler) in R to estimate unknown parameters with Markov Chain Monte Carlo (MCMC) methods in Bayesian framework. If you are interested in learning more about the theoretical knowledge of MCMC and Bayesian modeling, please check out courses "Probability and Statistics" (taught by Prof. Joseph Chang) and "Bayesian Statistics" (taught by Prof. Josh Warren). 
\
\
In the first section, we will introduce a simple jags model to help understand the syntax for rjags package.
\
\
In the second section, we will learn how to build up complicated models step by step (censor data). Before reading this section, you will need to familiar yourself with the RSV_timing_explain tutorials and a research article on the [relative RSV timing](https://doi.org/10.1111/irv.12965)
\
\
In the third section, we will introduce the concept of spatial autocorrelation and how to incorporate this in jags model.

@@

After we estimated the [RSV timing](https://doi.org/10.1111/irv.12965) using second derivative method (please check: RSV_timing_explain), we would like to know if some potential drivers contribute to the relative timing difference in RSV epidemic onsets before and during COVID-19 pandemic. (In reality, we manually coded the [spatial autocorrelated model](https://github.com/Gigi112/RSVtiming/blob/main/spatialstatis_best.R). As this requires taking Josh's classes, I will not go over it in the introduction tutorials.)

The model is given as: 
$$Y_i=\alpha_{0i}+\beta_1\overline{W_i}+\epsilon_i$$
$$\alpha_{0i}=\beta_0+\boldsymbol{x_i^T \lambda}+\theta_i$$

@@

where $Y_i$ is the relative timing of RSV epidemic onset with respect to Florida in 2021 in state $i$ (measured in weeks), and $\overline{w_i}$ is the average relative timing of RSV seasons in state $i$ compared to Florida from 2016-2019. The global intercept parameter $\beta_0$ and slope parameter $\beta_1$ describe the overall similarity between a typical year’s relative onset timing across all states and the relative timings observed in 2021. The observation-level random effect $\epsilon_i  ~ N(0,\sigma^2)$ accounts for residual variability in the data.
\
\
The state-specific intercepts, $\alpha_{0i}=\beta_0+\boldsymbol{x_i^T \lambda}+\theta_i$, represent the state-level similarity in relative timing for 2021 compared to previous seasons, where $x_i$ is the vector of covariates that could potentially explain differences in the relative timing of RSV between 2021 and previous years. The covariates include population density, average household size, and stringency index of non-pharmaceutical interventions against COVID-19. 
\
\
$\theta_i$ represents a spatially correlated random effect, which we will learn more in detail later.

## Jags model code without considering spatial autocorrelation

```{r}
Nospatial_model <- "model{
################ model ####################################
for(i in 1:n){
Y[i] ~ dnorm(mu[i], inv_var)
# for the observed timing (we log transform Y already)
 mu[i] <- alpha_0[i] + beta1*log_W[i]
 # log_W is the log transform average relative timing of RSV

alpha_0[i]= beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i]
# this is alpha_0i = beta0 + X_i*lambda (coviariates)
}

################ priors ####################################
beta0~dnorm(0,0.0001)
beta1~dnorm(0,0.0001)
lambda1~dnorm(0,0.0001)
lambda2~dnorm(0,0.0001)
lambda3~dnorm(0,0.0001)

inv_var ~ dgamma(0.01, 0.01)}"
```

## Load the data and run the jags model to sample posteriors

```{r, message=FALSE}
## loading data from the working directory
data_nospatial <- readRDS("./Nospatial_data.rds")

### Read in the model file using the jags.model function ##
non_spatial <- jags.model(textConnection(Nospatial_model),
                     data=data_nospatial,n.chains = 4)
```

```{r}
############ burn-in period ############
update(non_spatial,10000)

######## sample for posterior distribution ##########
non_spatial_sample <- coda.samples(non_spatial,
                            c("beta0","beta1",
                            "lambda1","lambda2",
                            "lambda3","inv_var"),
                             100000,thin=20)
```

## Summarize the posterior distribution

```{r}
summary(non_spatial_sample)
```

## Check for model convergence (gelman diagnostic)
**inv_var did not converge well**. This indicates that we should either prolong the iterations or re-parameterize the model. Using gelman.plot (next slides), we could tell that prolong iterations may help convergence.
```{r}
gelman.diag(non_spatial_sample)
```
## Gelman plot

```{r}
par(mfrow=c(3,2))
gelman.plot(non_spatial_sample)
```

## Trace plots
**note: inv_var is precision = the inverse of variance** 
```{r}
par(mfrow=c(3,2))
traceplot(non_spatial_sample)
```

## Density plot of the posterior distributions

```{r}
par(mfrow=c(3,2))
densplot(non_spatial_sample)
```

## Section 2.1: right censored survival analysis

In the early phase of an epidemic, we may not yet observe the onset point. For example, if you consider an onset threshold and the cases are still below the threshold, the epidemic has not taken off. In this case, we need to use right censored survival analysis. 
\
\
Click to know more about [right censored survival analysis](https://tinyheero.github.io/2016/05/12/survival-analysis.html)

## Model structure for right censored survival analysis

```{r, eval=FALSE}
censormodel <- "model{
for(i in 1:n){
is.censor[i] ~ dinterval(Z[i],t.cen)
# whether the data is censor and the censoring time

Z[i] ~ dlnorm(mu[i], inv_var)
# for the observed timing (we did not log transform Z)
# if we already log transform Z then we use dnorm

 mu[i] <- alpha_0[i] + beta1*log_W[i]
# log_W is the log transform average relative timing of RSV  
alpha_0[i]= beta0+x1[i]*lambda1+x2[i]*lambda2
# this is alpha_0i = beta0 + X_i*lambda (coviariates)
}
```

## Model structure for right censored survival analysis (continued)

```{r, eval=FALSE}
################ priors ####################################
beta0~dnorm(0,0.0001)
beta1~dnorm(0,0.0001)
lambda1~dnorm(0,0.0001)
lambda2~dnorm(0,0.0001)
inv_var ~ dgamma(0.01, 0.01)}"
```
\
\
Since the model running process is the same, I will skip the model fitting process in this tutorial (it takes time to knit the PDF). To run the model, you need to combine the two chunks of model structures and change the eval from False to True.

## Same process to sample posteriors of right censored model

```{r, message=FALSE, eval=FALSE}
## loading data from the working directory
data_censor <- readRDS("./censor_data.rds")

### Read in the model file using the jags.model function ##
censor_model_jags <- jags.model(textConnection(censormodel),
                     data=data_censor,n.chains = 4)

############ burn-in period ############
update(censor_model_jags,10000)

######## sample for posterior distribution ##########
censor_model_sample <- coda.samples(censor_model_jags,
                            c("beta0","beta1",
                            "lambda1","lambda2",
                            "lambda3","inv_var"),
                             100000,thin=20)
```

## Section 3: spatial autocorrelation
